Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules

### 2015-06-11 paper.tex
more writing

@@

%TODO: basic results

@@

@@

\begin{prop}\label{prop:set-of-reps}
  % There is a set of representatives for a basis of $\A(n)\mmod\A(n-k)$ (in the sense of the above note), all of which lie in $\L(k)$.
  Any element $t$ of $\A(n)$ may be written as $x + yz$ where $x\in \L(k)$, $y\in \A$, and $z\in\A(n-k)\setminus \F_2$ for any $k\geq 0$.  Equivalently, $\A(n)\subset \L(k) + \A(\Sq^i)_{i\leq k}$, where we take $\A(\Sq^i)_{i\leq k}$ to be the left ideal of $\A$ generated by $\{\Sq^i\}_{i\leq k}$, for any $k\geq 0$.
\end{prop}

Proposition~\ref{prop:set-of-reps} essentially says that there exists a set of representatives for a basis of $\A(n)\mmod\A(n-k)$, all of which lie in $\L(k)$.

\begin{proof}
  It suffices to prove the proposition on some basis for $\A(n)$.  Recall that in Wood's Z-basis of the Steenrod algebra, a basis for $\A(n)$ consists of all strings of squares selected from~\eqref{eq:woodz} and multiplied in the given order:
  \begin{equation} \label{eq:woodz}
    \Sq^{2^{n+1}-2^n}\Sq^{2^{n+1}-2^{n-1}}\cdots\Sq^{2^{n+1}-1} \Sq^{2^n-2^{n-1}}\cdots\Sq^{2^n-1}\cdots\Sq^2\Sq^3\Sq^1
  \end{equation}

  We induct on the number of squares selected.  The empty product is $1$, which is in $\L(k)$, so we are done.  So now suppose the proposition holds on all elements of the Wood Z-basis of length at most $l-1$, and let $t$ be an element of length $l$.  Then, for some $m=2^j-2^i$, $i<j\leq n+1$, we may write $t=\Sq^m\left(x+yz\right)$ where $x$, $y$, and $z$ are as in the statement of the proposition.

@@

  Write $x$ in admissible form as $\Sq^{i_1}\Sq^{i_2}\cdots\Sq^{i_k}$.  If $\Sq^m\Sq^{i_1}\Sq^{i_2}\cdots\Sq^{i_k}$ is already in admissible form, then we have that $2^ki_k\leq 2^{k-1}i_{k-1}\leq\cdots\leq2i_1\leq m<2^{n+1}$, so in particular, $i_k<2^{n-k+1}$, so $i_k\in\A(\Sq^i)_{i\leq k}$, so in fact the entire product $\Sq^mx$ is in $\A(\Sq^i)_{i\leq k}$.  Then we may apply the Adem relations at least once, to get
  \[\Sq^m\Sq^{i_1}\Sq^{i_2}\cdots\Sq^{i_k} = a_0\Sq^{m+i_1}\Sq^{i_2}\cdots\Sq^{i_k} + \sum_{1\leq j\leq \frac{m}2}a_j\Sq^{m+i_1-j}\Sq^j\Sq^{i_2}\cdots\Sq^{i_k};\]
  the first term we know is in $L(k)$.  For each of the remaining terms, if we may repeat the same process, ignoring the first factor; either the product is in $\A(\Sq^i)_{i\leq k}$, or it may be further reduced by the Adem relations.  Finally, if we apply the Adem relations $k$ times, the rightmost factor must be $\Sq^i$ for $i\leq \frac{m}{2^k}<2^{n-k+1}$, so similarly the product lies in the ideal $\A(\Sq^i)_{i\leq k}$.  Then $\Sq^mx$ may be written in the desired form, so we are done.
\end{proof}

@@

\begin{lem}\label{lem:li-vs}
  %TODO: is the claim here really that they are LI in the vector space over $\F2$?
  Taken as elements of $\Lkm$ as a vector space over $\F_2$, the set
  \[\{X_nX_{n-1}\cdots X_{n-k+1}\Sq^I\mid I=(2^{n+1}m_1,2^nm_2,\ldots,2^{n-k+2}m_k), m_1\geq\cdots\geq m_k\}\]
  is linearly independent.
\end{lem}
\begin{proof}
  Note that by the commutation relation we have shown, the element $s=X_nX_{n-1}\cdots X_{n-k+1}\Sq^I$ may be rewritten as
  \[X_n\Sq^{2^{n+1}m_1}X_{n-1}\Sq^{2^nm_2}\cdots X_{n-k+1}\Sq^{2^{n-k+2}m_k}.\]
  Let $M_{k,n}$ be the subspace of $\A$ which is spanned by monomials of the form $yz$ where $y\in L(k)$ and $z\in \A(n)$ with the degree of $z$ at least $1$.  If $b\in \A$, we will write $b+M_{k,n}$ for the subset of $\A$ consisting of elements of the form $b+m$ for $m\in M_{k,n}$.

@@

    %TODO: fix gap here: why isn't $X_n$ just in $M_{1, n-1}$ (may just need to show this out)?
  We proceed by induction on $k$ for fixed $n$.  If $k=1$, we know that $X_n\in M_{1,n-1}+\Sq^{n2^{n+1}+1}$, since the latter term is the only element of $L(1)$ in the right dimension.  Then since, as an $\A(n)$-module, $L(1)$ is periodic in degrees modulo $2^{n+1}$, $X_n\Sq^{2^{n+1}m}\in M_{1,n-1}+\Sq^{2^{n+1}(n+m)+1}$.

@@

  %TODO: show somewhere that the shortest terms are in fact the shortest, i.e., that all terms of $M_{k, n-k}$ are of length at least $k+1$.
  %TODO: make this argument clearer
  Then since the elements in question may be written in the given form, their shortest terms are unique and distinct, so the terms are linearly independent.
\end{proof}

@@

%TODO: I think we can't use the word basis here, we have to talk about filtrations, or "bases".
Lemma~\ref{lem:li-vs} suggests that the set
\[ B_{n,k} = \{Sq^I \mid I = (2^{n+1}m_1, 2^n m_2, \ldots, 2^{n-k+2}m_k), m_1 \geq \cdots \geq m_k \} \]
may be a basis for $\Lkm$ as an $\A(n)\mmod\A(n-k)$-module, through the following theorem.

\begin{thm}\label{thm:li-module}
  The set $B_{n,k}$ is linearly independent as elements of $\Lkm$ considered as an $\A(n)\mmod\A(n-k)$-module.
\end{thm}

@@

  %TODO: check that hopf algebra is the right criterion here, and that it's true
  First, consider the case where not all $b_i$ are in the same degree; let $b_j$ be one of the ones in the highest possible degree.  Since $\A(n)\mmod\A(n-k)$ is a Hopf algebra, there exists some $y_j$ in $\A(n)\mmod\A(n-k)$ such that $y_j x_j = z$.  Then for any $b_{j'}$ in a lower degree than $b_j$, the corresponding $x_{j'}$ will be in a higher degree than $x_j$.  Then $y_j x_j = 0$, so multiplying the entire relation by $y_j$ leaves a new relation amongst only the $b_i$ in the highest possible degree.

@@

Given this, it remains only to show that in some sense the basis $B_{n,k}$ has the right size.  Unfortunately, this is more difficult than it sounds.  If we take the full basis $B_{n,k}$ over $\Lkm$, the vector space of elements in each degree is infinite-dimensional, so the argument cannot proceed.  If we consider only the intersection of $B_{n,k}$ with $\L(k)$, this difficulty is removed, but another one presents itself: some classes are not spanned by the $B_{n,k}$.  While this issue is irrelevant to the desired final result, in that the periodic structure is still entirely encapsulated, with the exception of some initial conditions, a dimension-counting argument would require a careful accounting of these initial exceptions, which we found intractible in the general case.

@@

%TODO: discussion of small cases

>>>>

%TODO: more intro

@@

@@

\subsection{The Adem Relations}

Recall that the Steenrod squares satisfy the Adem relations:
\[ \Sq^i\Sq^j = \sum_{m=0}^{m/2}\binom{j-m-1}{i-2m}\Sq^{i+j-k}\Sq^k .\]
The binomial coefficient is taken modulo 2, and it is a standard result in combinatorics that it is nonzero if and only if for each bit of the binary representation of $i-2m$ that is 1, the corresponding bit of $j-m-1$ must also be 1.  We will use this fact extensively.

@@

\begin{lem}\label{lem:set-of-reps}
  % There is a set of representatives for a basis of $\A(n)\mmod\A(n-k)$ (in the sense of the above note), all of which lie in $\L(k)$.
  Any element $t$ of $\A(n)$ may be written as $x + yz$ where $x\in \L(k)$, $y\in \A$, and $z\in\A(n-k)\setminus \F_2$ for any $k\geq 0$.  Equivalently, $\A(n)\subset \L(k) + \A(\Sq^{2^i})_{i\leq n-k}$, where we take $\A(\Sq^{2^i})_{i\leq n-k}$ to be the left ideal of $\A$ generated by $\{\Sq^{2^i}\}_{i\leq n-k}$, for any $k\geq 0$.
\end{lem}

Proposition~\ref{lem:set-of-reps} essentially says that there exists a set of representatives for a basis of $\A(n)\mmod\A(n-k)$, all of which lie in $\L(k)$.

\begin{proof}
  It suffices to prove the proposition on some basis for $\A(n)$; we use the Wood $Z$ basis.  We induct on the number of squares selected to form the basis element.  The empty product is $1$, which is in $\L(k)$, so we are done.  So now suppose the proposition holds on all elements of the Wood Z-basis of length at most $l-1$, and let $t$ be an element of length $l$.  Then, for some $m=2^j-2^i$, $i<j\leq n+1$, we may write $t=\Sq^m\left(x+yz\right)$ where $x$, $y$, and $z$ are as in the statement of the proposition.

@@

  Write $x$ in admissible form as $\Sq^{i_1}\Sq^{i_2}\cdots\Sq^{i_k}$.  If $\Sq^m\Sq^{i_1}\Sq^{i_2}\cdots\Sq^{i_k}$ is already in admissible form, then we have that $2^ki_k\leq 2^{k-1}i_{k-1}\leq\cdots\leq2i_1\leq m<2^{n+1}$, so in particular, $i_k<2^{n-k+1}$, so $\Sq^{i_k}\in\A(\Sq^{2^i})_{i\leq n-k}$, so in fact the entire product $\Sq^mx$ is in $\A(\Sq^{2^i})_{i\leq n-k}$.  Then we may apply the Adem relations at least once, to get
  \begin{multline*}
  \Sq^m\Sq^{i_1}\Sq^{i_2}\cdots\Sq^{i_k} = a_0\Sq^{m+i_1}\Sq^{i_2}\cdots\Sq^{i_k} \\
                  + \sum_{1\leq j\leq \frac{m}2}a_j\Sq^{m+i_1-j}\Sq^j\Sq^{i_2}\cdots\Sq^{i_k};
  \end{multline*}
  the first term we know is in $L(k)$.  For each of the remaining terms, if we may repeat the same process, ignoring the first factor; either the product is in $\A(\Sq^{2^i})_{i\leq n-k}$, or it may be further reduced by the Adem relations.  Finally, if we apply the Adem relations $k$ times, the rightmost factor must be $\Sq^i$ for $i\leq \frac{m}{2^k}<2^{n-k+1}$, so similarly the product lies in the ideal $\A(\Sq^{2^i})_{i\leq n-k}$.  Then $\Sq^mx$ may be written in the desired form, so we are done.
\end{proof}

@@

\begin{lem}\label{lem:li-vs}
  Taken as elements of $\Lkm$ as a vector space over $\F_2$, the set
  \[\{X_nX_{n-1}\cdots X_{n-k+1}\Sq^I\mid I=(2^{n+1}m_1,2^nm_2,\ldots,2^{n-k+2}m_k), m_1\geq\cdots\geq m_k\}\]
  is linearly independent.
\end{lem}
\begin{proof}
  Note that by the commutation relation we have shown, the element $s=X_nX_{n-1}\cdots X_{n-k+1}\Sq^I$ may be rewritten as
  \[X_n\Sq^{2^{n+1}m_1}X_{n-1}\Sq^{2^nm_2}\cdots X_{n-k+1}\Sq^{2^{n-k+2}m_k}.\]
  Let $M_{k,n}$ be the subspace of $\A$ which is spanned by monomials of the form $yz$ where $y\in L(k)$ and $z\in \A(n)$ with the degree of $z$ at least $1$.  If $b\in \A$, we will write $b+M_{k,n}$ for the subset of $\A$ consisting of elements of the form $b+m$ for $m\in M_{k,n}$.

@@

    We proceed by induction on $k$ for fixed $n$.  If $k=1$, we know by Lemma~\ref{lem:set-of-reps} that $X_n\in M_{1,n-1}+\Sq^{n2^{n+1}+1}$; the latter term is the only element of $\L(1)$ in the right dimension, so we know it must be the value of $x$ from the lemma.  Then since, as an $\A(n)$-module, $L(1)$ is periodic in degrees modulo $2^{n+1}$, $X_n\Sq^{2^{n+1}m}\in M_{1,n-1}+\Sq^{2^{n+1}(n+m)+1}$.

@@

  %TODO: show somewhere that the shortest terms are in fact the shortest, i.e., that all terms of $M_{k, n-k}$ are of length at least $k+1$.
  Note that the terms 
\begin{equation}\Sq^{2^{n+1}(m_n+n)+1}\Sq^{2^n(m_{n-1}+n-1)+1}\cdots\Sq^{2^{n-k+2}(m_{n-k+1}+n-k+1)+1}\label{eq:b-term}\end{equation}
  are outside of $M_{k, n-k}$ since all elements of $\A(n-k)$ have degree smaller than the last factor of~\eqref{eq:b-term}.  Then since the elements in question may be written in the given form, we know that each has a unique term outside of $M_{k, n-k}$.  Those terms are distinct, so the terms are linearly independent.
\end{proof}

@@

Lemma~\ref{lem:li-vs} suggests that the set
\[ B_{n,k} = \{Sq^I \mid I = (2^{n+1}m_1, 2^n m_2, \ldots, 2^{n-k+2}m_k), m_1 \geq \cdots \geq m_k \} \]
may allow us to construct a filtration of $\Lkm$ with copies of $\A(n)\mmod\A(n-k)$ as the filtration quotients, through the following theorem.

\begin{thm}\label{thm:li-module}
  The set $B_{n,k}$ is linearly independent in $\Lkm$ with respect to coefficients taken from $\A(n)\mmod\A(n-k)$.
\end{thm}

@@

  First, consider the case where not all $b_i$ are in the same degree; let $b_j$ be one of the ones in the highest possible degree.  Since $\A(n)\mmod\A(n-k)$ is a Poincar\'e algebra, there exists some $y_j$ in $\A(n)\mmod\A(n-k)$ such that $y_j x_j = z$.  Then for any $b_{j'}$ in a lower degree than $b_j$, the corresponding $x_{j'}$ will be in a higher degree than $x_j$.  Then $y_j x_j = 0$, so multiplying the entire relation by $y_j$ leaves a new relation amongst only the $b_i$ in the highest possible degree.

@@

Given this, it remains only to show that in some sense the ``basis'' $B_{n,k}$ has the right size.  Unfortunately, this is more difficult than it sounds.  If we take the full $B_{n,k}$, the vector space of elements in each degree of $\Lkm$ is infinite-dimensional, so the argument cannot proceed.  If we consider only the intersection of $B_{n,k}$ with $\L(k)$, this difficulty is removed, but another one presents itself: some classes are not spanned by the $B_{n,k}$.  While these are irrelevant to the desired final result, in that the periodic structure is still entirely encapsulated, with the exception of some initial conditions, a dimension-counting argument would require a careful accounting of these initial exceptions, which we found intractible in the general case.

@@

Fortunately, in small cases the dimension-counting is tractable.  We begin by considering $\Lkm[1]$ as a module over some $\A(n)$.  The space $\Lkm[1]$ is particularly simple, since it is one-dimensional in each integral degree.

\begin{thm}
  As an $\A(n)$-module, $\Lkm[1]$ has a filtration with each filtration quotient consisting of a copy of $\A(n)\mmod\A(n-1)$.
\end{thm}
\begin{proof}
  We begin by showing that $\L(1)$ has such a filtration, with a single finite-dimensional base space of the filtration not isomorphic to $\A(n)\mmod\A(n-1)$.  The filtration quotients, in particular, are each a copy of $\A(n)\mmod\A(n-1)$ built on each element of $B_{n,1}\cap\L(1)$.

  These quotients are linearly independent by Theorem~\ref{thm:li-module}.  In addition, I claim they contain one class in each degree above $(n-1)2^{n+1}$.  In particular, consider some degree $d = m2^{n+1} - d'$, where $d' < 2^{n+1}$ and $m \geq n$.  Suppose that the binary representation of $d'$ has ones in the bits $k_1, k_2, \ldots, k_j$.  Then the class 
  \[\Sq^{2^{n+1}-2^{k_1}}\Sq^{2^{n+1}-2^{k_2}}\cdots\Sq^{2^{n+1}-2^{k_j}}\Sq^{(m-j)2^{n+1}} \]
  has is in the correct degree and is in the proposed filtration.  So the proposed filtration indeed spans all of $\L(1)$ with all quotients copies of $\A(n)\mmod\A(n-1)$ except possibly for some finite-dimensional base space containing classes in small degrees.

  Since the structure of $\Lkm[1]$ is periodic, this finite-dimensional base disappears if we extend the filtration to $\Lkm[1]$, so it has a filtration with all quotients consisting of a copy of $\A(n)\mmod\A(n-1)$ as desired.
\end{proof}

%TODO: pictures of the filtrations of L(1) and L(2)?

Next we consider the case of $\L(2)$.  Welcher's result shows that $\L(2)$ is free as an $\A(1)$-module.  As an $\A(2)$-module, we wish to show that $\Lkm[2]$ admits a filtration with quotients $\A(2)\mmod\A(0)$.  Again, only the dimension-counting remains.

\begin{thm}
  As an $\A(2)$-module, $\Lkm[2]$ has a filtration with each filtration quotient consisting of a copy of $\A(n)\mmod\A(n-1)$.
\end{thm}
\begin{proof}
  Again we begin by showing that $\L(2)$ has a filtration, this time with a single base space which we construct, and which is finite in each degree, and each other filtration quotient is isomorphic to $\A(2)\mmod\A(0)$, again built on elements of $B_{2,2}\cap \L(2)$.

  These quotients are linearly independent by Theorem~\ref{thm:li-module}.  In addition, I claim that none of the filtration quotients built on this basis span any of the following basis elements of $\L(2)$:
\begin{equation} \Sq^1, \Sq^3\Sq^1, \Sq^4\Sq^1, \Sq^5\Sq^1, \ldots. \label{eq:extra}\end{equation}
  This is easy to see from the Adem relations: clearly the only elements from $B_{2,2}\cap \L(2)$ that could be multiplied by elements of $\A(2)$ to get anything ending in $\Sq^1$ are the $\Sq^{8i}$, and it is easy to see from the Adem relations that any product which could create a term in~\eqref{eq:extra} and no others must involve multiplying by something ending in $\Sq^1$, i.e., by something which is zero in $\A(2)\mmod\A(0)$.  (Note that this is the point where using $\L(2)$ rather than $L(2)$ makes things much simpler: the set of excluded elements would be much more complicated in $L(2)$.)

  We now use Poincar\'e series to show that after adding those elements from~\eqref{eq:extra}, the claimed filtration quotients add up to have the right dimensionality in each degree, showing that the claimed filtration exists.  Recall that the Poincar\'e series of a graded vector space is the infinite series $v(x) = \sum_{i=0}^{\infty} x^i \dim V_i$.  By definition, the Poincar\'e series of $\L(2)$ is
  \[ l(x) = 1 + x + x^2 + 2x^3 + 2x^4 + 2x^5 + 3x^6 + \cdots = (1 + x + x^2 + \cdots)(1 + x^3 + x^6 + \cdots).\]
  Furthermore, from looking at the $Z$ basis, we can see that the Poincar\'e series of $\A(2)\mmod\A(0)$ is
  \[ a(x) = (1+x^4)(1+x^6)(1+x^7)(1+x^2)(1+x^3).\]
  The Poincar\'e series of the set $B_{2,2}\cap \L(2)$ is
  \[ b(x) = (1 + x^8 + x^{16} + \cdots)(1 + x^{12} + x^{24} + \cdots).\]
  Finally, the Poincar\'e series of the extra elements from~\eqref{eq:extra} is
  \[ c(x) = x + x^3 + x^4 + x^5 + \cdots.\]
  We wish to show that $l(x) = a(x)b(x) + c(x)$.  This is just a matter of algebra:
  \begin{align*}
    a(x)b(x) + c(x) &= (1+x^7)(1 + x^2 + x^4 + \cdots)(1 + x^3 + x^6 + \cdots) + c(x) \\
                    &= (1-x+x^2-\cdots+x^6)(1 + x + x^2 + \cdots)(1 + x^3 + x^6 + \cdots) \\
                    &\qquad + x + x^3(1 + x + x^2 + \cdots) \\
                    &= (1 + x + x^2 + \cdots)(1 - x + x^2 + x^3 + x^6 + x^9 + \cdots) + x \\
                    &= l(x) + x(x-1)(1 + x + x^2 + \cdots) + x \\
                    &= l(x) - x + x \\
    a(x)b(x) + c(x) &= l(x)
  \end{align*}
  Then since this filtration has linearly independent filtration quotients, and has the right total dimensionality, it must span the entirety of $\L(2)$.  As we continue backwards to $\Lkm[2]$, the excluded elements are not a part of the periodic structure, so they get filled in, and we get that $\Lkm[2]$ has a filtration consisting only of copies of $\A(2)\mmod\A(0)$.
\end{proof}

%TODO: acks

%TODO: bib
